An Extendable \proglang{Python} Implementation of Robust Optimisation Monte Carlo

### 2022-02-02 paper.tex
Add comments and some changes. Ask HP.

%% - \Abstract{} almost as usual
\Abstract{ Performing the inference in statistical models with
  intractable likelihood is challenging; most likelihood-free
  inference (LFI) methods encounter problems in accuracy and
  efficiency. In this paper, we present the implementation of Robust
  Optimisation Monte Carlo (ROMC) in the Python package ELFI. ROMC is
  a novel and efficient (highly-parallelizable) LFI technique that
  provides accurate weighted samples from the posterior.  Our
  implementation can be used in two ways. A scientist may use it as an
  out-of-the-box LFI algorithm. We provide an easy-to-use API
  harmonized with the rest of the ELFI framework; therefore, it is
  straightforward to compare it with all the methods included in
  ELFI. On the other direction, we have carefully split ROMC into
  isolated algorithmic steps. Therefore, we allow the researcher to
  experiment, replacing components of the algorithm with
  custom-designed subparts. In both scenarios, the algorithmic steps
  of ROMC run in a fully-parallelizable manner, exploiting all CPU
  cores. We also provide helpful functionalities for (i) inspecting
  the inference process and (ii) evaluating the approximated
  posterior. Finally, we test the robustness of our implementation on
  some typical LFI examples.

@@

Simulator-based models are particularly captivating due to the
modeling freedom they provide. In essence, a simulator-based model can
describe any data generating mechanism that can be written as a finite
set of algorithmic steps. This modeling freedom comes at a cost;
performing the inference, i.e., sample or evaluate the posterior
distribution, is challenging.

Optimization Monte Carlo (OMC) proposed by~\citep{Meeds2015} was a
novel LFI approach for approximating the posterior
distribution. Afterwards, \citep{Forneron2016} provided a similar
method under `reverse sampler'. The key idea is turning the stochastic
data-generating mechanism into a set of fully-parallelizable
deterministic optimization processes. In their work,
~\citep{Ikonomov2019}, located some critical failure modes of OMC, and
proposed Robust OMC (ROMC) an alternative version of OMC with the
appropriate improvements.

In this paper, we present an extendible and parallelizable
implementation of ROMC at the \proglang{Python} package
\proglang{Engine for Likelihood-Free inference (ELFI)}. We follow
specific principles for making the implementation extendible. As we
will describe analytically in the next chapter, ROMC can be understood
not only as a specific algorithm, but mainly as a framework for
obtaining weighted samples from the posterior; it defines some
algorithmic task that must be fulfilled, without enforcing which
specific algorithm must be used for the tasks. Therefore, a researcher
may adopt ROMC's backbone approach, but experiment with alternative
methods to solve the distinct tasks. We have designed ROMC for
faciliating such alterations. Finally, we have tested the accuracy and
the efficienct of our implementation on the LFI examples supported by
the \proglang{ELFI} package.

@@

% \clearpage
\section{Background}
We first give a short introduction to implicit parametric models and
likelihood-free inference. We then focus on the methods relevant for
this paper: first the earlier Optmisation Monte Carlo (OMC) and then
its recent robust generalisation, robust OMC, for which we provide the
software package as part of the Engine for likelihood-free inference
(ELFI), a larger inference platform that we briefly introduce at the
end of this background section.

@@

In this section, we present the implementation. At~\ref{subsec:ELFI}
we introduce the reader to the \proglang{Python} package \pkg{Engine
  for likelihood-free inference
  (ELFI)}. At~\ref{subsec:general_design} we explain the general
design that we follow. Then, we divide the implementation into three
parts; at~\ref{subsec:training} we present the training (fitting)
part, at~\ref{subsec:inference} the inference part and
at~\ref{subsec:evaluation} the evaluation part. Finally,
at~\ref{subsec:extensibility} we describe how a user may extend ROMC
with its custom modules.

@@

In this section, we test the implementation using the second-order
moving average (MA2) example, which is a standard model of
\pkg{ELFI}. We perform the inference using three different versions of
ROMC; (i) using a gradient-based optimizer, (ii) using the Bayesian
Optimization scheme and (iii) fitting a Neural Network as a surrogate
model. The later illustrates how to extend the implementation,
replacing part of ROMC with a user-defined component. Finally, we
measure the execution speed-up achieved by the parallelized version of
ROMC.

@@

In this paper, we presented the implementation details we followed for
developing the LFI method ROMC at the \proglang{ELFI} package. We paid
thorough attention to two specific use-case scenarios. Firstly, we
illustrate how a user may take advantage of our ready-to-use API for
solving its LFI problem. Secondly, we focus on the scenario where a
researcher wants to intervene and alter parts of the method. Our
implementation is designed to support this as well.

By all means, there are still open challenges for the future. Two
directions may enable ROMC to solve high-dimensional problems
efficiently. The first one is enabling ROMC's execution into a cluster
of computers. ROMC can be characterized as an \textit{embarrassingly
  parallel} workload; each optimization problem is an entirely
independent task. Therefore, supporting inference into a cluster of
computers can radically speed up the inference. The second one refers
to implementing the method in a framework that supports automatic
differentiation. Automatic differentiation is necessary for
efficiently solving optimization problems, especially in
high-dimensional parametric models.

>>>>

%% - \Abstract{} almost as usual
\Abstract{ Performing the inference in statistical models with
  intractable likelihood is challenging; most likelihood-free
  inference (LFI) methods encounter problems in accuracy and
  efficiency. In this paper, we present the implementation of Robust
  Optimisation Monte Carlo (ROMC) in the Python package ELFI. ROMC is
  a novel and efficient (highly-parallelizable) LFI technique that
  provides accurate weighted samples from the posterior.  Our
  implementation can be used in two ways. A scientist may use it as an
  out-of-the-box LFI algorithm. We provide an easy-to-use API
  harmonised with the rest of the ELFI framework; therefore, it is
  straightforward to compare it with all the methods included in
  ELFI. On the other direction, we have carefully split ROMC into
  isolated algorithmic steps. Therefore, we allow the researcher to
  experiment, replacing components of the algorithm with
  custom-designed subparts. In both scenarios, the algorithmic steps
  of ROMC run in a fully-parallelizable manner, exploiting all CPU
  cores. We also provide helpful functionalities for (i) inspecting
  the inference process and (ii) evaluating the approximated
  posterior. Finally, we test the robustness of our implementation on
  some typical LFI examples.

@@

Simulator-based models are particularly captivating due to the
modelling freedom they provide. In essence, a simulator-based model can
describe any data generating mechanism that can be written as a finite
set of algorithmic steps. This modelling freedom comes at a cost;
performing the inference, i.e., sample or evaluate the posterior
distribution, is challenging.

Optimization Monte Carlo (OMC) proposed by~\citep{Meeds2015} was a
novel LFI approach for approximating the posterior
distribution. Afterwards, \citep{Forneron2016} provided a similar
method under `reverse sampler'. The key idea is turning the stochastic
data-generating mechanism into a set of fully-parallelizable
deterministic optimisation processes. In their work,
~\citet{Ikonomov2019}, located some critical failure modes of OMC, and
proposed Robust OMC (ROMC) an alternative version of OMC with the
appropriate improvements.

In this paper, we present an extendible and parallelisable
implementation of ROMC at the \proglang{Python} package
\proglang{Engine for Likelihood-Free inference (ELFI)}. We follow
specific principles for making the implementation extendible. As we
will describe analytically in the next chapter, ROMC can be understood
not only as a specific algorithm, but mainly as a framework for
obtaining weighted samples from the posterior; it defines some
algorithmic task that must be fulfilled, without enforcing which
specific algorithm must be used for the tasks. Therefore, a researcher
may adopt ROMC's backbone approach, but experiment with alternative
methods to solve the distinct tasks. We have designed ROMC for
facilitating such alterations. Finally, we have tested the accuracy and
the efficiency of our implementation on the LFI examples supported by
the \proglang{ELFI} package.

@@

% \clearpage
\section{Background}
We first give a short introduction to implicit parametric models and
LFI. We then focus on the methods relevant for
this paper: first the earlier OMC and then
its recent robust generalisation, robust OMC, for which we provide the
software package as part of ELFI, a larger inference platform that we briefly introduce at the
end of this background section.

@@

In this section, we present the implementation. At~\ref{subsec:ELFI}
we introduce the reader to the \proglang{Python} package \pkg{ELFI}. At~\ref{subsec:general_design} we explain the general
design that we follow. Then, we divide the implementation into three
parts; at~\ref{subsec:training} we present the training (fitting)
part, at~\ref{subsec:inference} the inference part and
at~\ref{subsec:evaluation} the evaluation part. Finally,
at~\ref{subsec:extensibility} we describe how a user may extend ROMC
with its custom modules.

@@

In this section, we test the implementation using the second-order
moving average (MA2) example, which is one of the standard models of
\pkg{ELFI}. We perform the inference using three different versions of
ROMC; (i) using a gradient-based optimiser, (ii) using the Bayesian
Optimisation scheme and (iii) fitting a Neural Network as a surrogate
model. The later illustrates how to extend the implementation,
replacing part of ROMC with a user-defined component. Finally, we
measure the execution speed-up achieved by the parallelised version of
ROMC.

@@

In this paper, we presented the implementation details we followed for
developing the LFI method ROMC at the \pkg{ELFI} package. We paid
thorough attention to two specific use-case scenarios. Firstly, we
illustrate how a user may take advantage of our ready-to-use API for
solving its LFI problem. Secondly, we focus on the scenario where a
researcher wants to intervene and alter parts of the method. Our
implementation is designed to support this as well.

There are still open challenges for the left for future research. Two
directions may enable ROMC to solve high-dimensional problems
efficiently. The first one is enabling ROMC's execution into a cluster
of computers. ROMC can be characterized as an \textit{embarrassingly
  parallel} workload; each optimization problem is an entirely
independent task. Therefore, supporting inference into a cluster of
computers can radically speed up the inference. The second one refers
to implementing the method in a framework that supports automatic
differentiation. Automatic differentiation is necessary for
efficiently solving optimisation problems, especially in
high-dimensional parametric models.
